---
title: "Evolution Project - Sourcetracker2 Stacked Barplots"
output: html_notebook
author: "James Fellows Yates"
editor_options: 
  chunk_output_type: inline
---

# Introduction

For this project we want to do a small test to compare what happens
when including variations on the sources from the Evolution project.
In this context we want to see what is the difference of replacing
sediment with  a 'bone' source (as an archaeologically oral-microbiome
lacking source), and also including both at the same time.

Load libraries

```{r}
library(tidyverse)
```

# Preparation

Load data and select options

```{r}
## load data
 st2_raw_both <- read_tsv("../04-analysis/screening/sourcetracker.backup/otu_table_L6_1000_20190520/sink_predictions.txt")

 st2_meta <- read_tsv("../00-documentation.backup/02-calculus_microbiome-deep_evolution-individualscontrolssources_metadata_20190509.tsv")

colnames(st2_raw_both)[1] <- "sample"

colnames(st2_meta)[1] <- "sample"

```

Default ordering information

```{r}
env_ordering <- c("Howler_Monkey",
                                         "Gorilla_1",
                                         "Gorilla_2",
                                         "Gorilla_3",
                                         "Chimp_1",
                                         "Chimp_2",
                                         "Chimp_3",
                                         "Chimp_4",
                                         "Neanderthal",
                                         "PreagriculturalHuman_1",
                                         "PreagriculturalHuman_2",
                                         "PreantibioticHuman_1",
                                         "PreantibioticHuman_2",
                                         "ModernDayHuman_1",
                                         "ModernDayHuman_2",
                                         "ExtractionControl",
                                         "LibraryControl",
                                         "subPlaque",
                                         "supPlaque",
                                         "urbanGut",
                                         "ruralGut",
                                         "skin",
                                         "EnvironmentalControl",
                                         "sediment")
hostgenus_ordering <- c("Alouatta", "Gorilla", "Pan", "Homo", "Control")

```


# Data Clean and Function Definitions

For more intuitive plotting, we now need to manipulate the data for ordering
and colouring of the different sources. We can also hard-assign the colours 
for each source for consistency across each source.

```{r}
colnames(st2_raw_both) <- gsub("Z-", "", colnames(st2_raw_both))
colnames(st2_raw_both) <- gsub("Unknown", "unknown", colnames(st2_raw_both))

st2_meta <- st2_meta %>% 
  mutate(Env = as_factor(Env), 
         Env = fct_relevel(Env, env_ordering),
         Host_Genus = as_factor(Host_Genus),
         Host_Genus = fct_relevel(Host_Genus, hostgenus_ordering))

data_reorder <- function(data_raw, metadata, sample_order){
  ## Give: sink_predictions, metadata table and boolean for whether
  ## to order by sample name or proportion of subPlaque
  if (sample_order == TRUE) {
      data_mod <- data_raw %>% 
        arrange(-subPlaque) %>% 
        gather(source, proportion, 2:ncol(data_raw))
  } else {
      data_mod <- data_raw %>% 
      gather(source, proportion, 2:ncol(data_raw))
  }

    
  data_mod$source <- factor(data_mod$source, 
                                levels = rev(c("subPlaque", 
                                               "supPlaque", 
                                               "ruralGut", 
                                               "urbanGut", 
                                               "skin", 
                                               "EnvironmentalControl",
                                               "sediment",
                                               "unknown"))) 
  
  data_mod_meta <- left_join(data_mod, metadata) %>%
    arrange(Env)
  
  data_mod_meta$Env <- factor(data_mod_meta$Env, 
                              levels = env_ordering
                              )
  
  data_mod_meta <- data_mod_meta %>% 
    mutate(sample = factor(sample, unique(sample)))
}

source_colours <- c("#9e0142", "#d53e4f", "#f46d43", "#fdae61", "#abdda4", "#66c2a5", "#3288bd", "#5e4fa2")
names(source_colours) <- c("subPlaque", "supPlaque",  "ruralGut", "urbanGut", "skin", "EnvironmentalControl", "sediment", "unknown")

sourcetracker_bar_plotter <- function(input){
  ggplot(input, aes(sample, proportion, fill = source)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = source_colours) +
  xlab("Individuals") +
  ylab("Proportion") +
  labs(fill = 'Source') +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
        text = element_text(family = "Roboto"),
        legend.position = "bottom") +
  facet_grid(~Env, scales = "free", space = "free", switch = "y")
}
```

# Plotting

Then we can do our plotting.

```{r}
## For both
barplot_prefilter <- sourcetracker_bar_plotter(data_reorder(st2_raw_both, st2_meta, T))

barplot_prefilter

ggsave(plot = barplot_prefilter, 
       paste("sourcetracker_barplot_prefilter_",format(Sys.Date(), "%Y%m%d"),".pdf", sep = ""),
       "../04-analysis/screening/sourcetracker.backup/", 
       device = cairo_pdf, 
       width = 7, 
       height = 3.5,
       units = "in", 
       dpi = 600)

```

# Sample Filtering

Continuing with both sources together, we can now filter out samples 
which have a proportion of plaque that is less than either the sediment or 
skin component of the same sample.

Note, I have included 'gut' in the calculations. This is because it seems in
the chimps, there is a consistent level of a gut proportion in all samples 
which could be a true biological signal. Also, without considering it we lose
almost all samples (keeping only 4), despite CDFP plots showing that they are
well preserved. Furthermore, relatives to (organ protected) gut taxa are less 
likely to be taphonomic contaminants compared to skin.

```{r}
st2_filtering_calculations <- st2_raw_both %>% 
  rowwise() %>%
  mutate(total_plaquegut = sum(`subPlaque` + `supPlaque` + `ruralGut` + `urbanGut`)) %>%
  rowwise() %>%
  mutate(more_env = if_else(total_plaquegut < `sediment`, T, F)) %>%
  rowwise() %>%
  mutate(more_env = if_else(total_plaquegut < `skin`, T, more_env)) %>%
  rowwise() %>%
  mutate(more_env = if_else(total_plaquegut < unknown, T, more_env)) %>%
  rowwise() %>%
  mutate(more_env = if_else(total_plaquegut < EnvironmentalControl, T, more_env)) %>%
  right_join(st2_meta %>% select(sample, Env, Description, 
                                Host_General, Host_Common, Age), .)

write_tsv(st2_filtering_calculations, 
          paste("../00-documentation.backup/06-sourcetracker_filtering_results_", format(Sys.Date(), format = "%y%m%d"), ".tsv", sep = ""))

```

Save both results table and post-filtering barplot

```{r}

st2_filtered_both <- st2_filtering_calculations  %>% 
  filter(more_env == F) %>% 
  select(-more_env, -total_plaquegut, -Env, 
         -Description, -Host_General, -Host_Common, -Age)

barplot_postfilter <- sourcetracker_bar_plotter(data_reorder(st2_filtered_both, 
                                                             st2_meta, 
                                                             TRUE))

barplot_postfilter


ggsave(plot = barplot_postfilter, 
       paste("sourcetracker_barplot_postfilter_",format(Sys.Date(), "%Y%m%d"),".pdf", sep = ""),
       "../04-analysis/screening/sourcetracker.backup/", 
       device = cairo_pdf, 
       width = 7, 
       height = 3.5,
       units = "in", 
       dpi = 600)


```

# Post-filtering summary statistics

To save the samples to remove

And summary stats for the filtered results

```{r}

samples_to_retain <- st2_filtering_calculations %>% 
    select(sample, more_env) %>% 
    full_join(st2_meta, .) %>%
  filter(SourceSink == "sink")

```

```{r}

## Number of individuals per host genus population
samples_to_retain %>%
  select(sample, Env, more_env) %>%
  replace_na(replace = list(more_env = T)) %>%
  group_by(Env) %>% 
  summarise(No_Individuals = sum(!more_env)) %>% write_tsv(paste("../04-analysis/screening/sourcetracker.backup/post_filter_individual_hostgenuspopulation_counts_", format(Sys.Date(), "%Y%m%d"), ".tsv", sep = "")) %>% 
  print

## Number of individuals per host genus
samples_to_retain %>%
  select(sample, Host_Genus, more_env) %>%
  replace_na(replace = list(more_env = T)) %>%
  group_by(Host_Genus) %>% 
  summarise(No_Individuals = sum(!more_env)) %>%
  write_tsv(paste("../04-analysis/screening/sourcetracker.backup/post_filter_individual_hostgenus_counts_", format(Sys.Date(), "%Y%m%d"), ".tsv", sep = "")) %>% print

```


# Threshold Comparison
 
Without group assessment and purely based on proportion of plaque/gut

```{r}

lowest_proportion <- st2_filtering_calculations %>% 
  filter(more_env == F) %>%
  select(sample, Env, total_plaquegut) %>%
  pull(total_plaquegut) %>%
  min

st2_preservation_order <- st2_raw_both %>% 
  mutate(total_plaque = pmap(list(sub = subPlaque, sup = supPlaque, 
                               rural = ruralGut, urban = urbanGut),
                             function(sub, sup, rural, urban) {
                               sum(sub + sup + rural + urban)
         }) %>% 
           unlist) %>% 
  arrange(desc(total_plaque)) %>%
  mutate(sample = as_factor(sample)) %>%
  select(-total_plaque) %>%
  gather(source, proportion, 2:ncol(.)) %>% 
  mutate(source = as_factor(source),
         source = fct_relevel(source, rev(c("subPlaque", "supPlaque", 
                                            "ruralGut", "urbanGut", "skin", 
                                            "EnvironmentalControl", "sediment", 
                                            "unknown")))) %>%
  print()

abovebelow_plot <- ggplot(st2_preservation_order, aes(sample, proportion, fill = source)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = lowest_proportion) +
  scale_fill_manual(values = source_colours) +
  xlab("Individuals") +
  ylab("Proportion") +
  labs(fill = 'Source') +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
        text = element_text(family = "Roboto"),
        legend.position = "bottom")

abovebelow_plot

ggsave(plot = abovebelow_plot, 
       paste("sourcetracker_barplot_abovebelowthresholdcomparison_plaquegut_",format(Sys.Date(), "%Y%m%d"),".pdf", sep = ""),
       "../04-analysis/screening/sourcetracker.backup/", 
       device = cairo_pdf, 
       width = 7, 
       height = 3.5, 
       units = "in",
       dpi = 600)

```

Without group assessment and purely based on proportion of plaque only

```{r}

lowest_proportion <- st2_filtering_calculations %>% 
  filter(more_env == F) %>%
  rowwise() %>%
  mutate(total_plaque = sum(`subPlaque` + `supPlaque`)) %>%
  select(sample, Env, total_plaque) %>%
  pull(total_plaque) %>%
  min

st2_preservation_order <- st2_raw_both %>% 
  mutate(total_plaque = pmap(list(sub = subPlaque, sup = supPlaque),
                             function(sub, sup, rural, urban) {
                               sum(sub + sup)
         }) %>% 
           unlist) %>% 
  arrange(desc(total_plaque)) %>%
  mutate(sample = as_factor(sample)) %>%
  select(-total_plaque) %>%
  gather(source, proportion, 2:ncol(.)) %>% 
  mutate(source = as_factor(source),
         source = fct_relevel(source, rev(c("subPlaque", "supPlaque", 
                                            "ruralGut", "urbanGut", "skin", 
                                            "EnvironmentalControl", "sediment", 
                                            "unknown")))) %>%
  print()

abovebelow_plot <- ggplot(st2_preservation_order, aes(sample, proportion, fill = source)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = lowest_proportion) +
  scale_fill_manual(values = source_colours) +
  xlab("Individuals") +
  ylab("Proportion") +
  labs(fill = 'Source') +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
        text = element_text(family = "Roboto"),
        legend.position = "bottom")

abovebelow_plot

ggsave(plot = abovebelow_plot, 
       paste("sourcetracker_barplot_abovebelowthresholdcomparison_plaqueonly_",format(Sys.Date(), "%Y%m%d"),".pdf", sep = ""),
       "../04-analysis/screening/sourcetracker.backup/", 
       device = cairo_pdf, 
       width = 7, 
       height = 3.5, 
       units = "in",
       dpi = 600)

```

