---
title: "PhILR PCoA and PERMANOVA"
---

# Preamble

Following the introducory vignette: http://bioconductor.org/packages/release/bioc/vignettes/philr/inst/doc/philr-intro.html

Purpose of this notebook is to statistically test whether individuals from 
Homo fall distintctly based on whether they are from pre- or post-agricultural
time periods.

# Preparation

## For script conversion

As this notebook became more and more parameterised, I decided to make it
easy to convert into a script. But for the script, we also need to define
input arguments for the options in chunk 1.

```{r purl = FALSE}
## Name options CHANGE HERE
db <- "nt" ## nt or refseq
tax_level <- "genus" ## genus or species
sources <- "noSources" ## "withSources" "noSources" 
controls <- "noControls" ## "withControls" "noControls"
bad_samples <- "out" ## in or out
sample_filter <- "withinvariation" # onepcburnin, twocburnin, fivepcburnin, tenpcburnin, withinvariation or none
script <- F
minsupp_multiplier <- 1
minsupp_threshold <- 0.01 * minsupp_multiplier
###########################################
```

```{r  fig.height=3.5, fig.width=7}
args = commandArgs(trailingOnly = TRUE)

if (args[1] == "" | args[1] == "-h" | args[1] == "--help") {
 cat("Usage: 017-PhILR_PCoA_20XXXXXX_script.R <db> <tax_level> <sources> <controls> <bad_samples> <sample_filter> <minsupp_multiplier>\n")
 cat("db: nt or refseq \n")
 cat("tax_level: genus or species \n") 
 cat("sources: noSources or withSources \n")
 cat("controls: noControls or withControls \n")
 cat("bad_samples: in or out \n")
 cat("sample_filter: sourcetracker, onepcburnin, twocburnin, fivepcburnin, tenpcburnin, withinvariation or none \n")
 cat("minsupp_multiplier: number to multiply foundation 0.01% minimum support")
 stop()
} else if (length(args) == 7) {
 db <- args[1]
 tax_level <- args[2]
 sources <- args[3]
 controls <- args[4]
 bad_samples <- args[5]
 sample_filter <- args[6]
 minsupp_multiplier <- as.numeric(args[7])
 script <- T
}

cat(args)

minsupp_multiplier <- as.numeric(minsupp_multiplier)
minsupp_threshold <- 0.01 * minsupp_multiplier
 
```


## Data Loading and Cleaning

Load libraries

```{r  fig.height=3.5, fig.width=7}
library(tidyverse) ## for general data cleaning
library(taxize) ## for NCBI taxonomic info collection
library(ape) ## for tree manipulation
library(phyloseq) ## for data format as input for PhILR
library(philr) ## for data transform
library(vegan) ## for statistical testing
library(ggtree) ## for tree visualisation
library(patchwork) ## for further visualisation assistance
library(broom) ## for saving stats summary tables
library(plotly) ## for interactive plots
library(pairwiseAdonis) # for PERMANOVA mucltiple correction testing
library(fpc) ## for hclustering stuff
library(clues) ## for hclustering stuff
```

Load already generated data from MEGAN and metadata. We also need to export the 
same data with as a tree from MEGAN with the option: file > export > tree.

```{r  fig.height=3.5, fig.width=7}

## The tree related to the OTU table
if (tax_level == "genus" & db == "nt") {
 otu_tree <- read.tree("../04-analysis/screening/megan.backup/Evolution-Comparison_20190401_nt_prokaryotes_genus.nwk")
} else if (tax_level == "species" & db == "nt") {
 otu_tree <- read.tree("../04-analysis/screening/megan.backup/Evolution-Comparison_20190401_nt_prokaryotes_species.nwk")
} else if (tax_level == "genus" & db == "refseq") {
 otu_tree <- read.tree("../04-analysis/screening/megan.backup/Evolution-Comparison_20190410_refseq_prokaryotes_genus.nwk")
} else if (tax_level == "species" & db == "refseq") {
  otu_tree <- read.tree("../04-analysis/screening/megan.backup/Evolution-Comparison_20190410_refseq_prokaryotes_species.nwk")
}


## OTU tables
if (tax_level == "genus" & db == "nt") {
 otu_table <- read_tsv("../04-analysis/screening/megan.backup/Evolution-Comparison_MEGAN_20190401-ex_absolute_genus_prokaryotes_summarised_nt.txt")
} else if (tax_level == "species" & db == "nt") {
 otu_table <- read_tsv("../04-analysis/screening/megan.backup/Evolution-Comparison_MEGAN_20190401-ex_absolute_species_prokaryotes_summarised_nt.txt")
} else if (tax_level == "genus" & db == "refseq") {
 otu_table <- read_tsv("../04-analysis/screening/megan.backup/Evolution-Comparison_MEGAN_20190410-ex_absolute_genus_prokaryotes_summarised_refseq.txt")
} else if (tax_level == "species" & db == "refseq") {
 otu_table <- read_tsv("../04-analysis/screening/megan.backup/Evolution-Comparison_MEGAN_20190410-ex_absolute_species_all_summarised_refseq.txt")
}


## Predicted contaminant taxa to remove
if (tax_level == "genus" & db == "nt") {
 taxa_to_remove <- read_tsv("../04-analysis/screening/decontam.backup/decontam_taxa_to_remove_megan_nt_genus_combined_0.99_190411.tsv")
} else if (tax_level == "species" & db == "nt") {
 taxa_to_remove <- read_tsv("../04-analysis/screening/decontam.backup/decontam_taxa_to_remove_megan_nt_species_combined_0.99_190411.tsv")
} else if (tax_level == "genus" & db == "refseq") {
 taxa_to_remove <- read_tsv("../04-analysis/screening/decontam.backup/decontam_taxa_to_remove_megan_refseq_genus_combined_0.99_190411.tsv")
} else if (tax_level == "species" & db == "refseq") {
 taxa_to_remove <- read_tsv("../04-analysis/screening/decontam.backup/decontam_taxa_to_remove_megan_refseq_species_combined_0.99_190411.tsv")
}

## Metadata
raw_metadata <- read_tsv("../00-documentation.backup/02-calculus_microbiome-deep_evolution-individualscontrolssources_metadata_20190523.tsv")

## Bad samples to remove

if (sample_filter == "sourcetracker") {
 samples_to_remove <- read_tsv("../04-analysis/screening/sourcetracker.backup/sourcetracker_filtering_results_190509.tsv") %>% 
  filter(more_env == T)
} else if (db == "nt" && sample_filter == "onepcburnin") {
 samples_to_remove <- read_tsv("home/fellows../04-analysis/screening/cumulative_decay.backup/cumulativeproportiondecay_burninfilter1pc_nt_fractionOralThreshold_50_20190509.tsv") %>% 
  filter(more_env == T)
} else if (db == "nt" && sample_filter == "twopcburnin") {
 samples_to_remove <- read_tsv("home/fellows../04-analysis/screening/cumulative_decay.backup/cumulativeproportiondecay_burninfilter2pc_nt_fractionOralThreshold_50_20190509.tsv") %>% 
  filter(more_env == T)
} else if (db == "nt" && sample_filter == "fivepcburnin") {
 samples_to_remove <- read_tsv("../04-analysis/screening/cumulative_decay.backup/cumulativeproportiondecay_burninfilter5pc_nt_fractionOralThreshold_50_20190509.tsv") %>% 
  filter(more_env == T)
} else if (db == "nt" && sample_filter == "tenpcburnin") {
 samples_to_remove <- read_tsv("../04-analysis/screening/cumulative_decay.backup/cumulativeproportiondecay_burninfilter10pc_nt_fractionOralThreshold_50_20190509.tsv") %>% 
  filter(more_env == T)
} else if (db == "nt" && sample_filter == "withinvariation") {
 samples_to_remove <- read_tsv("../04-analysis/screening/cumulative_decay.backup/cumulativeproportiondecay_burninwithinfluctuationSDvariation_nt_fractionOralThreshold_50_20190509.tsv") %>% 
  filter(more_env == T)
} else if (db == "refseq" && sample_filter == "onepcburnin") {
 samples_to_remove <- read_tsv("home/fellows../04-analysis/screening/cumulative_decay.backup/cumulativeproportiondecay_burninfilter1pc_refseq_fractionOralThreshold_65_20190509.tsv") %>% 
  filter(more_env == T)
} else if (db == "refseq" && sample_filter == "twopcburnin") {
 samples_to_remove <- read_tsv("home/fellows../04-analysis/screening/cumulative_decay.backup/cumulativeproportiondecay_burninfilter2pc_refseq_fractionOralThreshold_65_20190509.tsv") %>% 
  filter(more_env == T)
} else if (db == "refseq" && sample_filter == "fivepcburnin") {
 samples_to_remove <- read_tsv("../04-analysis/screening/cumulative_decay.backup/cumulativeproportiondecay_burninfilter5pc_refseq_fractionOralThreshold_65_20190509.tsv") %>% 
  filter(more_env == T)
} else if (db == "refseq" && sample_filter == "tenpcburnin") {
 samples_to_remove <- read_tsv("../04-analysis/screening/cumulative_decay.backup/cumulativeproportiondecay_burninfilter10pc_refseq_fractionOralThreshold_65_20190509.tsv") %>% 
  filter(more_env == T)
} else if (db == "refseq" && sample_filter == "withinvariation") {
 samples_to_remove <- read_tsv("../04-analysis/screening/cumulative_decay.backup/cumulativeproportiondecay_burninwithinfluctuationSDvariation_refseq_fractionOralThreshold_65_20190509.tsv") %>% 
  filter(more_env == T)
}



```

Clean up to remove samples not required and then remove any OTUs that
now have no counts. Also remove OTUs that are likely lab contaminants.

Here we will also move all but ancient humans.

```{r  fig.height=3.5, fig.width=7}
data_cleaner <- function(x) {
 colnames(x) <- gsub("_S.*_L.*_R1_.*.fastq.combined.fq.prefixed.extractunmapped.bam","", colnames(x))
 colnames(x) <- gsub("_S.*_L00.*_R1_.*.fastq.merged.prefixed.hg19unmapped", "", colnames(x))
 colnames(x) <- gsub("_S.*_L00.*_R1_.*.fastq.extractunmapped.bam", "", colnames(x))
 colnames(x) <- gsub("_S.*_L.*_R1_.*.fastq.merged", "", colnames(x))
 colnames(x) <- gsub(".prefixed.hg19unmapped", "", colnames(x))
 colnames(x) <- gsub("_S0_L003_R1_001.sorted.bam.unmapped", "", colnames(x))
 colnames(x)[1] <- "Taxon"
 return(x)
}

## Remove col cruft of Metadata
raw_metadata <- rename(raw_metadata, Individual = `#SampleID`)

## Remove bad sources from OTU table
if (bad_samples == "in") {
 otu_table <- otu_table %>% 
 data_cleaner
} else if (bad_samples == "out") {
 otu_table <- otu_table %>% 
  data_cleaner %>% 
  dplyr::select(-one_of(samples_to_remove %>% 
                         left_join(raw_metadata, 
                                   by = c("sample" = "Individual")) %>%
                         select(sample, SourceSink, Sample_or_Control) %>% 
                         filter(SourceSink == "sink", 
                                Sample_or_Control == "Sample") %>% 
                          pull(sample)) 
 )
}
 


## Conditional filtering out of sources and/or Controls
if (sources == "withSources") {
 NA
} else if (sources == "noSources") {
 otu_table <- otu_table %>% 
 dplyr::select(Taxon, one_of(filter(raw_metadata, SourceSink == "sink") %>% 
             pull(Individual)))
}

if (controls == "withControls") {
 NA
} else if (controls == "noControls") {
 otu_table <- otu_table %>% 
 dplyr::select(Taxon, one_of(filter(raw_metadata, 
                                    Sample_or_Control == "Sample") %>% 
             pull(Individual)), 
             contains("ARS"))
}


## Filter taxa not passing min support threshold 
if (db == "nt") {
 otu_table <- otu_table %>% 
  gather(Individual, Value, 2:ncol(.)) %>% 
  left_join(select(raw_metadata, Individual, Min_Support_Reads_Threshold_MALT)) %>%
  mutate(Threshold = Min_Support_Reads_Threshold_MALT * minsupp_multiplier) %>%
  mutate(Threshold = as.numeric(Threshold)) %>%
  mutate(Filter_Passed = if_else(Value >= Threshold, 1, 0)) %>% 
  filter(Filter_Passed == 1) %>%
  select(Taxon, Individual, Value) %>%
  spread(Individual, Value, fill = 0)
} else if (db == "refseq") {
 otu_table <- otu_table %>% 
  gather(Individual, Value, 2:ncol(.)) %>% 
  left_join(select(raw_metadata, Individual, Min_Support_Reads_Threshold_MALT_refseq)) %>%
  mutate(Threshold = Min_Support_Reads_Threshold_MALT_refseq * minsupp_multiplier) %>%
  mutate(Threshold = as.numeric(Threshold)) %>%
  mutate(Filter_Passed = if_else(Value >= Threshold, 1, 0)) %>% 
  filter(Filter_Passed == 1) %>%
  select(Taxon, Individual, Value) %>%
  spread(Individual, Value, fill = 0)
}

## Remove ancient humans
keep_list <- raw_metadata %>% filter(Host_General %in% c("Neanderthal", "PreagriculturalHuman", "PreantibioticHuman", "ModernDayHuman")) %>% pull(Individual) %>% intersect(colnames(otu_table))
names(keep_list) <- keep_list

otu_table <- otu_table %>% select(Taxon, keep_list)

## Convert to matrix
otu_matrix <- as.matrix(dplyr::select(otu_table, -Taxon))
rownames(otu_matrix) <- otu_table$Taxon

## Remove any taxa that were unique to the bad samples
pos_otus <- rowSums(otu_matrix)
pos_otus <- pos_otus[pos_otus != 0] 

## Remove lab contaminants
otu_matrix <- subset(otu_matrix, !rownames(otu_matrix) %in% (taxa_to_remove %>% pull))
otu_matrix_final <- subset(otu_matrix, rownames(otu_matrix) %in% names(pos_otus))

rownames(otu_matrix_final) <- gsub(" ", "_", rownames(otu_matrix_final))
```

## Make Taxonomy Table

We also need to make our taxonomy table at species level. This takes a long 
to run, so I've generated ones in the past and re-load them when needed.

```{r eval = FALSE}
taxonomy_to_tibble <- function(x) {
 y <- gsub("\\[", "", x)
 y <- gsub("\\]", "", y)
 y <- gsub("\\(.*\\)", "", y)
 
 ttt_uid <- taxize::get_uid(y, db = "ncbi") ## inspired by myTAI::taxonomy()
 ttt_result <- taxize::classification(ttt_uid)[[1]] ## inspired by myTAI::taxonomy()
 
 if ( is.na(ttt_result)[1] ) {
  ttt_out <- x  
 } else {
  ttt_out <- as_tibble(ttt_result %>% 
   mutate(taxon = paste(x)) %>%
   filter(rank %in% c("superkingdom", "kingdom", "phylum", "class", "order", "family", "genus", "species")) %>%
   dplyr::select(name, rank, taxon) %>%
   spread(rank, name))
 }
 return(ttt_out)
  Sys.sleep(2)
}

# taxonomy_summary <- rownames(otu_matrix_final) %>% map(., .f = taxonomy_to_tibble)

taxonomy_summary <- tibble(taxon = character(), 
              superkingdom = character(),
              kingdom = character(),
              phylum = character(),
              class = character(), 
              order = character(),
              family = character(), 
              genus = character(),
              species = character()
              )

n <- 0
tot <- nrow(otu_matrix_final)
fail_taxa <- c()

## Temp due to curl timeout, change number beginning øf range depending on
## where n got to at time out
otu_matrix_final_sub <- otu_matrix_final[903:nrow(otu_matrix_final),]

## If time out, replace otu_matrix_final with otu_matrix_final_sub and re-run
for(i in rownames(otu_matrix_final_sub)) {
 print(paste(i," - ", format(round((n / tot) * 100), nsmall = 2), "%", sep = ""))
 n <- n + 1
 taxonomy_temp <-taxonomy_to_tibble(i)
 
 if (length(taxonomy_temp) = 1) {
  fail_taxa <-append(fail_taxa, i)
 } else {
   taxonomy_summary <- bind_rows(taxonomy_summary, taxonomy_temp)
 }
}

taxonomy_summary <- taxonomy_summary %>% distinct()

```

And save

```{r eval = FALSE}
save(taxonomy_summary, file = paste("../04-analysis/screening/philr_dietary.backup/0-evolution-philr_taxonomytable", "_", db,"_", tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".robj", sep = ""))
```

For the failed taxa, manually look up paths and add taxonomic paths and add them
and save.

```{r eval = FALSE}
## To check for the ones still missing
missing <- setdiff(rownames(otu_matrix_final), rownames(taxonomy_table_final))

out <- tibble(taxon = character(), 
              superkingdom = character(),
              kingdom = character(),
              phylum = character(),
              class = character(), 
              order = character(),
              family = character(), 
              genus = character(),
              species = character()
              )

n <- 0
tot <- length(missing)
fail_taxa <- c()

for(i in missing[1:length(missing)]) {
 print(paste(i," - ", format(round((n / tot) * 100), nsmall = 2), "%", sep = ""))
 n <- n + 1
 taxonomy_temp <- taxonomy_to_tibble(i)
 
 if (length(taxonomy_temp) == 1) {
  fail_taxa <- append(fail_taxa, i)
 } else {
   out <- bind_rows(out, taxonomy_temp)
 }
}

taxonomy_summary <- bind_rows(taxonomy_summary, out) %>% distinct()


## Some manual additions
temp <- read_tsv("~/Downloads/Genus_extra.csv")
taxonomy_summary <- bind_rows(taxonomy_summary, temp) %>% distinct()

save(taxonomy_summary, file = "../04-analysis/screening/philr_dietary.backup/0-evolution-philr_taxonomytable_withsources_withControls_all_20190404.robj")
```

Load a previously made taxonomy table

```{r  fig.height=3.5, fig.width=7}
if (db == "nt") {
 load("../04-analysis/screening/philr_dietary.backup/0-evolution-philr_taxonomytable_withsources_withblanks_all_20190404.robj")
} else if (db == "refseq") {
 load("../04-analysis/screening/philr_dietary.backup/0-evolution-philr_taxonomytable_refseq_species_withSources_withControls_20190115.robj")
}

taxonomy_table <- as.data.frame(taxonomy_summary[2:ncol(taxonomy_summary)])
rownames(taxonomy_table) <- taxonomy_summary$taxon 
taxonomy_table <- as.matrix.data.frame(taxonomy_table)


## To make sure OTU names match tree, which in import replaced spaces with 
## undercores
#rownames(taxonomy_table) <- gsub(" ", "_", rownames(taxonomy_table))

## Filter for genus level
if (tax_level == "genus") {
 taxonomy_table_final <- taxonomy_table[,c("superkingdom", 
                                           "kingdom", 
                                           "phylum", 
                                           "class", 
                                           "order", 
                                           "family", 
                                           "genus")] %>% 
  as.data.frame.matrix
 rownames(taxonomy_table_final) <- c()
 taxonomy_table_final <- subset(taxonomy_table_final, 
                                genus %in% gsub("_", 
                                                " ", 
                                                rownames(otu_matrix_final))) %>% 
   unique()
 taxonomy_table_final <- unique(taxonomy_table_final)
 row.names(taxonomy_table_final) <- taxonomy_table_final[,"genus"]
 taxonomy_table_final <- as.matrix(taxonomy_table_final)
 taxonomy_table_final <- unique(taxonomy_table_final)

 } else if (tax_level == "species") {
 taxonomy_table_final <- taxonomy_table
}
```


## Check Tree Format Valid

As need a purely bifurcating tree (thus do not allow polytomy), and need a
rooted tree we should check this with:

```{r  fig.height=3.5, fig.width=7}
is.rooted(otu_tree)
is.binary.tree(otu_tree)
```

If we don't have a binary tree, solve polytomies with 

```{r  fig.height=3.5, fig.width=7}
otu_tree <- multi2di(otu_tree)
is.binary.tree(otu_tree)
```

Or if this doesn't work we can remove singletons (as in single-intermediate 
nodes between a leaf and a parent node) with

```{r  fig.height=3.5, fig.width=7}
otu_tree <- collapse.singles(otu_tree) # from http://blog.phytools.org/2018/05/when-phylogeny-fails-isbinary-but-is.html
is.binary.tree(otu_tree)

```

## Prepare Metdata

Prepare sample metadata

```{r  fig.height=3.5, fig.width=7}

raw_metadata <- raw_metadata %>%
  mutate(Region = if_else(Location %in% c("Morroco", "Ethiopia", "South_Africa"), "Africa", "Europe"))

meta_data <- as.data.frame(raw_metadata[2:ncol(raw_metadata)], 
              row.names = raw_metadata$Individual, 
              colnames = col.names(raw_metadata[2:col(raw_metadata),]))
rownames(meta_data) <- raw_metadata$Individual

```


# Clustering Analysis

## PhILR Transform

### Make PhyloSeq object

Make our Phyloseq object.

```{r  fig.height=3.5, fig.width=7}
ps_data <- phyloseq(otu_table(otu_matrix_final, taxa_are_rows = TRUE), tax_table(taxonomy_table_final), phy_tree(otu_tree), sample_data(meta_data))
```

### Zero Replacement

We also need to a run zero removal procedure as the *LR methods require
positive numbers. Here just doing pseudocount (note might be flawed - see Fodor 
paper), but here just following PhILR tutorial. Have used zCompositions
in the past but that was with random guessing of parameters (or using
recommended defaults)

```{r  fig.height=3.5, fig.width=7}
ps_data <- transform_sample_counts(ps_data, function(x) x + 1)
```

### Final Checks

We fix the tree nodes with and something else from the PhILR tutorial I don't 
fully understand with name balancing and transpose to sample row/taxa column
composition convention.

```{r  fig.height=3.5, fig.width=7}
phy_tree(ps_data) <- makeNodeLabel(phy_tree(ps_data), method = "number", prefix = 'n')
name.balance(phy_tree(ps_data), tax_table(ps_data), 'n1')

otu.table <- t(otu_table(ps_data))
tree <- phy_tree(ps_data)

```

Some last checks

```{r  fig.height=3.5, fig.width=7}
otu.table <- t(otu_table(ps_data))
tree <- phy_tree(ps_data)
metadata <- sample_data(ps_data)

## Add new region column thing
metadata$Region <- if_else(metadata$Location %in% c("Morroco", "Ethiopia", "South_Africa"), "Africa", "Europe")

tax <- tax_table(ps_data)

otu.table[1:2,1:2] # OTU Table
tree
head(metadata,2)
```

### Run PhILR

Now run PhILR (hopefully)... here goes...

```{r  fig.height=3.5, fig.width=7}
ps_data.philr <- philr(otu.table, 
            tree,
            part.weights = "enorm.x.gm.counts",
            ilr.weights = "uniform")

ps_data.philr[1:5,1:5]
```

And we can now do ordination within the PhILR space with 
Euclidean distances

```{r  fig.height=3.5, fig.width=7}
ps_data.dist <- dist(ps_data.philr, method = "euclidean")
ps_data.pcoa <- ordinate(ps_data, 'PCoA', distance = ps_data.dist)
```

### Plot PhILR

Pre-plotting formatting

#### Aesethetics

```{r  fig.height=3.5, fig.width=7}
env_colours <- c("#1f78b4", 
                 "#6a3d9a", 
                 "#6a3d9a", 
                 "#6a3d9a", 
                 "#33a02c", 
                 "#33a02c", 
                 "#33a02c", 
                 "#33a02c", 
                 "#e31a1c", 
                 "#ff7f00", 
                 "#ff7f00", 
                 "#ff7f00", 
                 "#ff7f00", 
                 "#ff7f00", 
                 "#ff7f00", 
                 "#d9d9d9", 
                 "#d9d9d9",
                 "#d9d9d9", 
                 "#d9d9d9", 
                 "#d9d9d9", 
                 "#d9d9d9", 
                 "#d9d9d9", 
                 "#d9d9d9",
                 "#d9d9d9")

names(env_colours) <- c("Howler_Monkey", 
           "Gorilla_1", 
           "Gorilla_2", 
           "Gorilla_3", 
           "Chimp_1", 
           "Chimp_2", 
           "Chimp_3", 
           "Chimp_4",
           "Neanderthal", 
           "PreagriculturalHuman_1", 
           "PreagriculturalHuman_2", 
           "PreantibioticHuman_1", 
           "PreantibioticHuman_2", 
           "ModernDayHuman_1", 
           "ModernDayHuman_2", 
           "ExtractionControl", 
           "LibraryControl", 
           "ruralGut", 
           "urbanGut",
           "sediment", 
           "skin", 
           "subPlaque",
           "supPlaque",
           "EnvironmentalControl"
           )

 env_shapes <- c(8,
                0,
                1,
                2,
                0,
                1,
                2,
                5,
                11,
                0,
                12,
                1,
                10,
                2,
                6,
                10,
                13,
                7,
                8,
                14,
                3,
                4,
                12,
                0)

 
 names(env_shapes) <- c("Howler_Monkey", 
           "Gorilla_1", 
           "Gorilla_2", 
           "Gorilla_3", 
           "Chimp_1", 
           "Chimp_2", 
           "Chimp_3", 
           "Chimp_4",
           "Neanderthal", 
           "PreagriculturalHuman_1", 
           "PreagriculturalHuman_2", 
           "PreantibioticHuman_1", 
           "PreantibioticHuman_2", 
           "ModernDayHuman_1", 
           "ModernDayHuman_2", 
           "ExtractionControl", 
           "LibraryControl", 
           "ruralGut", 
           "sediment", 
           "skin", 
           "subPlaque",
           "supPlaque", 
           "urbanGut",
           "EnvironmentalControl"
          )
 
common_colours <- c(Alouatta = "#1f78b4", Gorilla = "#6a3d9a", Pan = "#33a02c", 
          `Homo (Neanderthal)` = "#ff7f00", 
          `Homo (Modern Human)` = "#ff7f00", ExtractionControl = "#d9d9d9", 
          LibraryControl = "#d9d9d9", Plaque = "#d9d9d9", Gut = "#d9d9d9", 
          Skin = "#d9d9d9", Sediment = "#d9d9d9", EnvironmentalControl = "#d9d9d9")

common_shapes <- c(Alouatta = 8, Gorilla = 0, Pan = 1, 
          `Homo (Neanderthal)` = 2, `Homo (Modern Human)` = 6,
          ExtractionControl = 10, LibraryControl = 13, Plaque = 9, 
          Gut = 4, Skin = 14, Sediment = 7, EnvironmentalControl = 12)

 general_colours <- c(Neanderthal = "#e31a1c", PreagriculturalHuman = "#d95f0e", PreantibioticHuman = "#ff7f00", ModernDayHuman = "#fdbb84")
 general_shapes <- c(Neanderthal = 11, PreagriculturalHuman = 0, PreantibioticHuman = 1, ModernDayHuman = 2)
 
 
 
```


And plot with host populations

```{r  fig.height=3.5, fig.width=5}

## Convert to tidy data and add metadata
ps_data.pcoa_tib <- as_tibble(ps_data.pcoa$vectors, rownames = "Individual") %>%
 left_join(as_tibble(metadata, rownames = "Individual"))

ps_data.pcoa_tib$Env <- factor(ps_data.pcoa_tib$Env, levels = c("Howler_Monkey", 
           "Gorilla_1", 
           "Gorilla_2", 
           "Gorilla_3", 
           "Chimp_1", 
           "Chimp_2", 
           "Chimp_3", 
           "Chimp_4",
           "Neanderthal", 
           "PreagriculturalHuman_1", 
           "PreagriculturalHuman_2", 
           "PreantibioticHuman_1", 
           "PreantibioticHuman_2", 
           "ModernDayHuman_1", 
           "ModernDayHuman_2", 
           "ExtractionControl", 
           "LibraryControl", 
           "ruralGut", 
           "sediment",
           "skin", 
           "subPlaque",
           "supPlaque", 
           "urbanGut",
           "EnvironmentalControl")
           )

## Extract percentage variation
percentage_data <- round(100 * ps_data.pcoa$values[1] / sum(ps_data.pcoa$values[1]), 2)
rownames(percentage_data) <- rownames(ps_data.pcoa$value)

## Colour/Shapes by sample type and source/sink
pcoa_plot_hostpops_12 <- ggplot(ps_data.pcoa_tib, aes(Axis.1, Axis.2, colour = Host_General, shape = Host_General, text = Individual)) +
 geom_point(size = 2, stroke = 0.5) +
 scale_colour_manual(values = general_colours) +
 scale_shape_manual(values = general_shapes) +
 xlab(paste("Axis 1 (", percentage_data[1,], "%)", sep = "")) +
 ylab(paste("Axis 2 (", percentage_data[2,], "%)", sep = "")) +
 theme_minimal(base_family = "Roboto", base_size = 7) +
 labs(shape = "Host Population", colour = "Host Population") +
  theme(legend.position = "bottom")

if (script == F) {
 pcoa_plot_hostpops_12
}

ggsave(paste("01a-philr_pcoa_malt_euclidean_axis1axis2_group_", db , "_", tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".pdf", sep = ""), plot = pcoa_plot_hostpops_12, "../04-analysis/screening/philr_dietary.backup/", device = cairo_pdf, width = 3.5, height = 3.5, units = "in", dpi = 600)

## Axis 3 and 2
pcoa_plot_hostpops_32 <- ggplot(ps_data.pcoa_tib, aes(Axis.3, Axis.2, colour = Host_General, shape = Host_General)) +
 geom_point(size = 2, stroke = 0.5) +
 scale_colour_manual(values = general_colours) +
 scale_shape_manual(values = general_shapes) +
 xlab(paste("Axis 3 (", percentage_data[3,], "%)", sep = "")) +
 ylab(paste("Axis 2 (", percentage_data[2,], "%)", sep = "")) +
 theme_minimal(base_family = "Roboto", base_size = 7) +
 labs(shape = "Host Population", colour = "Host Population") +
  theme(legend.position = "bottom")


if (script == F) {
pcoa_plot_hostpops_32
}

ggsave(paste("01b-philr_pcoa_malt_euclidean_axis3axis2_group_", db , "_", tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".pdf", sep = ""), plot = pcoa_plot_hostpops_32, "../04-analysis/screening/philr_dietary.backup/", device = cairo_pdf, width = 3,5, height = 3.5, units = "in", dpi = 600)


```

Colour by Region


```{r}
## Colour/Shapes by sample type and source/sink
pcoa_plot_hostregion_12 <- ggplot(ps_data.pcoa_tib, aes(Axis.1, Axis.2, colour = Region, shape = Host_General, text = Individual)) +
 geom_point(size = 2) +
 scale_shape_manual(values = general_shapes) +
 xlab(paste("Axis 1 (", percentage_data[1,], "%)", sep = "")) +
 ylab(paste("Axis 2 (", percentage_data[2,], "%)", sep = "")) +
 theme_minimal(base_family = "Roboto", base_size = 7) +
 labs(shape = "Host Population", colour = "Host Population") +
 theme(legend.position = "bottom") +
 guides(shape = F, colour = guide_legend(title = "Region"))


if (script == F) {
 pcoa_plot_hostregion_12
}

ggsave(paste("01a-philr_pcoa_malt_euclidean_axis1axis2_region_", db , "_", tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".pdf", sep = ""), plot = pcoa_plot_hostregion_12, "../04-analysis/screening/philr_dietary.backup/", device = cairo_pdf, width = 3.5, height = 3.5, units = "in", dpi = 600)


## Axis 3 and 2
pcoa_plot_hostregion_32 <- ggplot(ps_data.pcoa_tib, aes(Axis.3, Axis.2, colour = Region, shape = Host_General)) +
 geom_point(size = 2) +
 scale_shape_manual(values = general_shapes) +
 xlab(paste("Axis 3 (", percentage_data[3,], "%)", sep = "")) +
 ylab(paste("Axis 2 (", percentage_data[2,], "%)", sep = "")) +
 theme_minimal(base_family = "Roboto", base_size = 7) +
 labs(shape = "Host Population", colour = "Host Population") +
 theme(legend.position = "bottom")


if (script == F) {
pcoa_plot_hostregion_32
}

ggsave(paste("01b-philr_pcoa_malt_euclidean_axis3axis2_region_", db , "_", tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".pdf", sep = ""), plot = pcoa_plot_hostregion_32, "../04-analysis/screening/philr_dietary.backup/", device = cairo_pdf, width = 3.5, height = 3.5, units = "in", dpi = 600)

```
### Combined Group and Region

```{r}
dietary_pcoa_comparison <- pcoa_plot_hostpops_12 + pcoa_plot_hostregion_12

dietary_pcoa_comparison

ggsave(paste("11-philr_pcoa_malt_euclidean_axis1axis2_populationVSregion_", db , "_", tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".pdf", sep = ""), plot = dietary_pcoa_comparison, "../04-analysis/screening/philr_dietary.backup/", device = cairo_pdf, width = 7, height = 3.5, units = "in", dpi = 600)

ggsave(paste("11-philr_pcoa_malt_euclidean_axis1axis2_populationVSregion_", db , "_", tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".png", sep = ""), plot = dietary_pcoa_comparison, "../04-analysis/screening/philr_dietary.backup/", device = "png", width = 7, height = 3.5, units = "in", dpi = 600)
```



## Hierarchical Clustering

Moving away from the PhILR Tutorial, what happens if we run hierarchical 
cluster analysis instead, using the distance matrix made in the previous
method.

```{r, fig.width = 7, fig.height = 7}
## Cluster
ps_data.philr_hclust <- hclust(ps_data.dist, method = "average")

## Convert hclust to phylo object for plotting
ps_data.philr_clust_phylo <- as.phylo(ps_data.philr_hclust)

## Plot
ps_data.philr_clust_dendro_diet <- ggtree(as.phylo(ps_data.philr_clust_phylo), layout = "circular", branch.length = 'none', right = TRUE, ladderize = TRUE) %<+%
 raw_metadata + 
 geom_tiplab2(size = 2, hjust = -.2) +
 geom_tippoint(aes(shape = Host_General, colour = Host_General), stroke = 1.1, size = 3) +
 scale_shape_manual(values = general_shapes) +
 scale_color_manual(values = general_colours) +
 theme_tree() +
 theme(text = element_text(family = "Roboto", size = 7), legend.position = "bottom")

if (script == F) {
ps_data.philr_clust_dendro_diet
}

ggsave(paste("04a-philr_pcoa_malt_euclidean_hclustwardD2_hostgroup_", db , "_",tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".pdf", sep = ""), plot = ps_data.philr_clust_dendro_diet, "../04-analysis/screening/philr_dietary.backup/", device = cairo_pdf, width = 7, height = 7, units = "in", dpi = 600)

ggsave(paste("04a-philr_pcoa_malt_euclidean_hclustwardD2_hostgroup_", db , "_",tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".png", sep = ""), plot = ps_data.philr_clust_dendro_diet, "../04-analysis/screening/philr_dietary.backup/", device = "png", width = 7, height = 7, units = "in", dpi = 600)


## By region
ps_data.philr_clust_dendro_region <- ggtree(as.phylo(ps_data.philr_clust_phylo), layout = "circular", branch.length = 'none', right = TRUE, ladderize = TRUE) %<+%
 raw_metadata + 
 geom_tiplab2(size = 2, hjust = -.2) +
 geom_tippoint(aes(shape = Host_General, colour = Region), stroke = 1.1, size = 3) +
 scale_colour_brewer(palette = "Set1") +
 scale_shape_manual(values = general_shapes) +
 theme_tree() +
 theme(text = element_text(family = "Roboto", size = 7), legend.position = "bottom")

if (script == F) {
ps_data.philr_clust_dendro_region
}

ggsave(paste("04b-philr_pcoa_malt_euclidean_hclustwardD2_hostregion_", db , "_",tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".pdf", sep = ""), plot = ps_data.philr_clust_dendro_region, "../04-analysis/screening/philr_dietary.backup/", device = cairo_pdf, width = 7, height = 7, units = "in", dpi = 600)

ggsave(paste("04b-philr_pcoa_malt_euclidean_hclustwardD2_hostregion_", db , "_",tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".pdf", sep = ""), plot = ps_data.philr_clust_dendro_region, "../04-analysis/screening/philr_dietary.backup/", device = "png", width = 7, height = 7, units = "in", dpi = 600)
```

Attempt at cluster validation. The rand index looks through a pairwise matrix
to see in how many disagreements occur between the elements clusters
assignment of two clustering results. In otherwords, it compares one element
of one result to another element of the other result. Do they both have the 
same relationship (in terms of cluster assignment) or not?. You then get the 
fraction of disagreements over all comparisons, and subtract from one.

A value of 0 shows no pattern of relationship, whereas a value of 1 shows
an exact matches between the two clustering results.

Additionally, the adjustedRandIndex encorporates the possibility of 'chance' 
that two elements are classed in the same cluster, even if there is no 
correspondence between the two.

We can also look at the jaccard index, which is intersection over union, i.e.
the fraction of element agreements overall all comparisons. A value of 1 
represents and exact match whereas 0 have total disagreement.

References:
https://stackoverflow.com/questions/29520137/find-true-positive-true-negative-etc-in-case-of-clustering-location-data
https://davetang.org/muse/2017/09/21/the-rand-index/
https://davetang.org/muse/2017/09/21/adjusted-rand-index/
https://davetang.org/muse/2017/09/28/rand-index-versus-adjusted-rand-index/
https://stackoverflow.com/questions/42418773/how-can-we-interpret-negative-adjusted-rand-index

```{r}
expect_cluster <- metadata$Host_General
expect_cluster <- case_when(expect_cluster == "ModernDayHuman" ~ 1,
                            expect_cluster == "Neanderthal" ~ 2,
                            expect_cluster == "PreagriculturalHuman" ~ 2,
                            expect_cluster == "PreantibioticHuman" ~ 3)
names(expect_cluster) <- metadata$individual_id


hc_clusters <- cutree(ps_data.philr_hclust, 4)

cluster_comparison <- adjustedRand(expect_cluster, hc_clusters)

cluster_comparison

write_tsv(cluster_comparison %>% enframe(name = "Validation Metric", value = "Value"), paste0("../04-analysis/screening/philr_dietary.backup/05-philr_pcoa_malt_euclidean_hclust_externalvalidation_", db , "_",tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".tsv", sep = ""))

```

In both cases (HA for Hubert/Arabie version, Jaccard for Jaccard) we see that  
the agreements are closer to 0 than 1, suggesting that the clustering of these
individuals based on presumed dietary affiliation do not agree - when 
considering oral microbiome composition.

# Multivariate Analysis

Next we want to calculate if the groups by host are truly distinct. For
this people typically use PERMANOVA based on a distance matrix used to visualise
the groups. It is nice as it is semi-parametric (i.e. doesn't need normal
distributions), yet still allows downstream typical downstream statistical
tests.

Howver, we first need to check some assumptions related to it as described in
Anderson 2017 [(Wiley StatsRef: Statistics Reference Online) ](https://onlinelibrary.wiley.com/doi/full/10.1002/9781118445112.stat07841).
We can assume we have a good dissimilarity measure, as we have followed current
best practises for CoDa data and also are using the 'simplest' distance measure
which is Euclidean.

## Dispersion Homogenity Test

However, the second assumption is that there is homogenity of dispersions among
groups (or beta-dispersion). While the method is robust when this is applied
on balanced designs (i.e. same sample sizes per group), it is not robust
when you have different numbers of sample sizes (for example: 
http://thebiobucket.blogspot.com/2011/04/assumptions-for-permanova-with-adonis.html, 
or originally published Anderson and Walsh (2013 in Ecological Monographs 83(4))). 

Anderson thus recommends using the 'PERMDISP' test which is implemented as 
`betadisper()` in R's `vegan` library (originally published in Anderson et al. 
(2006) Biometrics 62(1) 245-253)

First we need to make a vector for the grouping of the sample, with the order 
matching that in the distance matrix.

```{r  fig.height=3.5, fig.width=7}
metadata_subset <- raw_metadata %>% 
 filter(Individual %in% rownames(otu.table)) %>% 
 dplyr::select(Individual, Host_General, Region)

ind_groups <- metadata_subset %>% pull(Host_General)
region <- metadata_subset %>% pull(Region)
```

Now we can run the `betadisper()` function to get the average distances from
the centroid

```{r  fig.height=3.5, fig.width=7}
ps_data.betadisper <- betadisper(ps_data.dist, ind_groups, type = "centroid")
ps_data.betadisper

## to save
sink(paste("../04-analysis/screening/philr_dietary.backup/06-betadisper_homogenity_", db , "_",tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".txt", sep = ""))
ps_data.betadisper
sink(file = NULL)

```

To check that we are violating the assumption of homogenity of 
variances (i.e. the amount beta-dispersion does not vary between each group), 
we can check this with an ANOVA.

```{r  fig.height=3.5, fig.width=7}
## to save

ps_data.betadisper_anova <- anova(ps_data.betadisper) %>% tidy

ps_data.betadisper_anova

write_tsv(ps_data.betadisper_anova, paste("../04-analysis/screening/philr_dietary.backup/07-betadisper_homogenity_anova_", db , "_", tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".txt", sep = ""))

```

Here, see we do have a borderline similarity in the amount of dispersion
between each group.

We can double-double check this with an alternative method suggested by the
`vegan` library developers, were we can use a permutation test.

```{r  fig.height=3.5, fig.width=7}
ps_data.betadisper_permutest <- permutest(ps_data.betadisper , pairwise = TRUE)
ps_data.betadisper_permutest

# to save
sink(paste("../04-analysis/screening/philr_dietary.backup/08-betadisper_homogenity_permutest_", db , "_", tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".txt", sep = ""))

ps_data.betadisper_permutest
sink(file = NULL)
```

For species and genus level, the F and P value suggest that the variances in the 
dispersions are indeed significantly different (i.e. unlikely to have been 
picked from the same distribution by chance).

According to the `vegan` help page on the `betadisper` function we can plot 
this with

```{r  fig.height=3.5, fig.width=7}

betadisper_scatterplot <- plot(ps_data.betadisper, main = 'PCoA Beta Dispersion of Groups', xlab = "Axis 1", ylab = "Axis 2", cex.lab = 0.7, cex.axis = 0.7, cex.main = 0.7, cex.sub = 0.7)

if (script == F) {
 betadisper_scatterplot
}

## Note: can't #ggsave plot()

genus_colours <- c(Alouatta = "#F7756C", Gorilla = "#A2A500", Pan = "#00BF7D", 
          Homo = "#E76BF2", Control = "#6a3d9a", Plaque = "#fdbf6f", 
          Gut = "#a6cee3", Skin = "#fb9a99", Sediment = "#b2df8a")


betadisper_boxplot <- tibble(ps_data.betadisper$distances, rownames = names(ps_data.betadisper$distances)) %>% 
 rename(Individual = rownames, Distance_to_centroid = `ps_data.betadisper$distances`) %>% 
 left_join(raw_metadata) %>% 
 rename(Host = Host_General) %>%
 ggplot(aes(Host, Distance_to_centroid, fill = Host)) + 
 geom_boxplot() + 
 xlab("Host Group") +
 ylab("Distance to Centroid") +
 scale_fill_manual(values = genus_colours) +
 theme_minimal(base_size = 7, base_family = "Roboto") +
 labs(fill = "Host Group") +
 theme(axis.text.x = element_text(angle = 30))

if (script == F) {
 betadisper_boxplot
}
ggsave(paste("09-philr_betadisper_malt_boxplot_hostgenus_", db , "_",tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".pdf", sep = ""), plot = betadisper_boxplot, "../04-analysis/screening/philr_dietary.backup/", device = cairo_pdf, width = 3.5, height = 3.5, units = "in", dpi = 600)

```

Where we can explicitly see large variations in dispersion.

## PERMANOVA

The PERMANOVA null hypothesis is that
under the assumption of 'exchangeability of the sample units among the groups'
(as in if you randomly re-assign a sample to a different group), the centroids 
(X~Y mean) among the groups is equvalient for all groups. Thus, if the null 
hypothesis is true then any observered differences would be the same as if you 
randomly allocated sample units to the groups. In contrast, if it is false, 
then the observed differences is actually from different centroids between 
the groups.

```{r  fig.height=3.5, fig.width=7}
permanova_factors <- factor(ind_groups)
original_result <- adonis(ps_data.dist ~ ind_groups) %>% print

## to save
sink(paste("../04-analysis/screening/philr_dietary.backup/10a-permanova_initial_hostgenuscentroids_", db , "_", tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".txt", sep = ""))
 original_result
sink(file = NULL)

## Store values for later
original_f <- original_result$aov.tab$F.Model[1]
original_p <- original_result$aov.tab$`Pr(>F)`[1]

```

The results here would suggest that we do a extremely borderly significantly 
different centroid of each group, suggesting that they are _maybe_ distinct. 
However, we violate of assumption of homogenity between the beta-dispersion of 
each group. This is possibly due to us having different sample sizes.

## PERMANOVA with Multiple Correction Testing 

We also need to perform multiple-testing correction as we are using the same
group over and over in this test, if I understand correctly.

```{r}
original_result_posthoc <- pairwise.adonis(ps_data.dist, ind_groups)

original_result_posthoc <- original_result_posthoc %>% as_tibble()
  
original_result_posthoc

write_tsv(original_result_posthoc, 
          paste("../04-analysis/screening/philr_dietary.backup/10b-permanova_initial_hostgenuscentroids_multiplecorrectiontesting_", db , "_", tax_level, "_", sources, "_", controls, "_badsamples", bad_samples, "_", sample_filter, "_", minsupp_threshold, "_" , format(Sys.Date(), "%Y%m%d"),".txt", sep = ""))



```

Here we see that there is only a single significantly different group which is
between Neandertals and Preantibiotic humans.

## Script generation

Use the following to generate a new fast script.

```{r purl = F}
knitr::purl("../02-scripts.backup/017-PERMANOVA_HomoOnly_Dietary_20190916.Rmd", ../02-scripts.backup/017-PERMANOVA_HomoOnly_Dietary_20190916_script.R", documentation = 2)
```

