---
title: "Core Microbiome - Database comparison"
output: html_notebook
---

## Preparation

This notebook shows describes the consistency between the core microbiome 
assignments between the two databases.

Load Libraries

```{r}
library(tidyverse)
library(ggalluvial)
library(scales)
library(ggbeeswarm)
library(data.table)
library(patchwork)
```

Load the data

```{r}
data_raw <- Sys.glob("../04-analysis/screening/presenceabsence_intersection.backup") %>%
  list.files("08a-*", full.names = T) %>% 
  enframe(name = NULL, value = "file_name") %>%
  mutate(file_contents = map(file_name, ~suppressMessages(read_tsv(.x))),
         Database = map(file_name, ~str_split(.x, "_")[[1]][7] %>% str_replace("maltdb", "") %>% as.character) %>% unlist,
         Min_Supp = map(file_name, ~str_split(.x, "_")[[1]][8]) %>% unlist,
         Inds_Frac = map(file_name, ~str_split(.x, "_")[[1]][9] %>% str_replace("fracinds", "") %>% as.numeric) %>% unlist,
         Pops_Frac = map(file_name, ~str_split(.x, "_")[[1]][10] %>% str_replace("fracpops", "") %>% as.numeric) %>% unlist,
         Single_Pops_Dropped = map(file_name, ~str_split(.x, "_")[[1]][11] %>% str_replace("singleindpopsdropped", "") %>% as.logical) %>% unlist,
         Grouping = map(file_name, ~str_split(.x, "_")[[1]][12]) %>% unlist,
         Level = map(file_name, ~str_split(.x, "_")[[1]][13]) %>% unlist,
         Date = map(file_name, ~str_split(.x, "_")[[1]][14]) %>% unlist %>% str_replace(".tsv", "")) %>%
  select(-file_name) %>%
  filter(Date == "20190902",
         !Min_Supp %in% c(0.08, 0.09) ) %>%
  unnest()

data_raw <- Sys.glob("../04-analysis/screening/presenceabsence_intersection.backup") %>%
  list.files("08b-*", full.names = T) %>% 
  enframe(name = NULL, value = "file_name") %>%
  mutate(file_contents = map(file_name, ~suppressMessages(read_tsv(.x))),
         Database = map(file_name, ~str_split(.x, "_")[[1]][7] %>% str_replace("maltdb", "") %>% as.character) %>% unlist,
         Min_Supp = map(file_name, ~str_split(.x, "_")[[1]][8]) %>% unlist,
         Inds_Frac = map(file_name, ~str_split(.x, "_")[[1]][9] %>% str_replace("fracinds", "") %>% as.numeric) %>% unlist,
         Pops_Frac = map(file_name, ~str_split(.x, "_")[[1]][10] %>% str_replace("fracpops", "") %>% as.numeric) %>% unlist,
         Single_Pops_Dropped = map(file_name, ~str_split(.x, "_")[[1]][11] %>% str_replace("singleindpopsdropped", "") %>% as.logical) %>% unlist,
         Grouping = map(file_name, ~str_split(.x, "_")[[1]][12]) %>% unlist,
         Level = map(file_name, ~str_split(.x, "_")[[1]][13]) %>% unlist,
         Date = map(file_name, ~str_split(.x, "_")[[1]][14]) %>% unlist %>% str_replace(".tsv", "")) %>%
  select(-file_name) %>%
  filter(Date == "20190902") %>%
  unnest() %>%
  bind_rows(data_raw) %>%
  filter(Min_Supp == 0.07 & Tax_Level == "genus" | Min_Supp == 0.04 & Tax_Level == "species",
         !Single_Pops_Dropped)
```

## General Database Comparison

Now spread to compare between the database

```{r}
data_match <- data_raw %>% distinct() %>% 
  spread(Database, Combination) %>% 
  select(Min_Supp, Tax_Level, Taxon, nt, refseq) %>% 
  mutate(Match = nt == refseq)


## Save for publication

data_match %>% 
  replace_na(list(nt = "DidNotPassPrevalenceThreshold", refseq = "DidNotPassPrevalenceThreshold", Match = "DidNotPassPrevalenceThreshold")) %>%
  write_tsv(paste0("../04-analysis/screening/presenceabsence_intersection.backup/99-coremicrobiome_presenceabsence_databasecomparison_table_", format(Sys.Date(), "%Y%m%d"), ".tsv"))
```

To summarise

```{r}
data_match %>% 
  group_by(Tax_Level, Match) %>% 
  summarise(`No. Matches` = n()) %>%
  group_by(Tax_Level) %>%
  mutate(total = sum(`No. Matches`))
```

We see above half of each category (assuming a taxon is found in both) is 
supported between the two databases.

We can check this using `nt` as our reference

```{r}
## Get the original number of taxa for each combination

data_combs <- data_raw %>% 
  group_by(Database, Tax_Level, Combination) %>% 
  summarise(`No. Taxa per Combination` = n())

data_corroboration <- data_match %>% 
  gather(Database, Combination, nt, refseq) %>% 
  filter(Match) %>%
  distinct() %>%
  group_by(Database, Tax_Level, Combination) %>% 
  summarise(`No. Taxa Corroborated` = n()) %>%
  arrange(Tax_Level,  desc(`No. Taxa Corroborated`)) %>% 
  left_join(data_combs) %>%
  mutate(`Combination Corroborated (%)` = (`No. Taxa Corroborated` / `No. Taxa per Combination` * 100)) %>%
  arrange(Tax_Level,  desc(`Combination Corroborated (%)`))

data_corroboration %>% 
  filter(Combination %in% c("Alouatta:Gorilla:Pan:Homo", "Gorilla:Pan:Homo", "Pan:Homo")) %>%
  arrange(Tax_Level, Database, desc(`Combination Corroborated (%)`))

```

Here we see that that when looking at only our host phylogeny
expected combinations (which we also use for the core), that at genus
level all genus-level >= 50% of the combinations are aggreed upon between
the two databases. There is less support at species level, with all combinations
are >= 60%.

## How are taxa reassigned?

We can investigate how the reassignment occurs (when a taxon is assigned between
two of the host phylogeny core combination  - not if a taxon is  reassigned to
a non-host phylogeny combination)

```{r}
data_match_reduced <- data_raw %>% 
  distinct() %>% 
  filter(Combination %in% c("Alouatta:Gorilla:Pan:Homo", "Gorilla:Pan:Homo", "Pan:Homo")) %>%
  spread(Database, Combination) %>% 
  select(Min_Supp, Tax_Level, Taxon, nt, refseq) %>% 
  mutate(Match = nt == refseq) %>%
  filter(!Match) %>%
  print()
```

In 3/4 cases, the reassignment is in the nt database having a taxon
being assigned to a Gorilla:... combination, whereas in RefSeq it is
the same combination but without Gorilla. The other is _Campylobacter gracilis_
being considered core to all anthropoids in nt, and just Hominins in Refseq.


```{r}
data_match_reduced <- data_raw %>% 
  distinct() %>% 
  filter(Combination %in% c("Alouatta:Gorilla:Pan:Homo", "Gorilla:Pan:Homo", "Pan:Homo")) %>%
  spread(Database, Combination) %>% 
  select(Min_Supp, Tax_Level, Taxon, nt, refseq) %>% 
  mutate(Match = nt == refseq) %>%
  filter(is.na(Match)) %>%
  print()
```

All other cases, it appears the taxa (mostly at species level), the differences
between the combination assignments rae likely due to the given taxon not 
passing prevalence or abundance thresholds in the other database.
