---
title: "Cumulative Percent Decay Filtering Results Summary"
output: html_notebook
---

# Introduction

The purpose of this notebook is to summaries the results of the output of
the cumulative percent decay plots for low preservation samples detection.

# Infrastructure

Set project directory

```{r}
proj_dir <- "/home/fellows/projects1/microbiome_calculus/evolution/"
```

## Libraries

```{r}
library(tidyverse)
```

# Data Wrangling

Load cumulative decay plots

```{r}
cumperdec <- Sys.glob(paste0(proj_dir, "04-analysis/screening/cumulative_decay.backup")) %>%
  list.files("*.tsv", full.names = T) %>%
  enframe(name = NULL, value = "file_path") %>%
  filter(grepl("burninwithinfluctuation", file_path),
         grepl("20190509", file_path)) %>%
  mutate(file_contents = map(file_path, ~read_tsv(.x, col_type = cols())),
         database = map(file_path, ~str_split(.x, "_")[[1]] %>% head(5) %>% tail(1)) %>% unlist,
         threshold = map(file_path, ~str_split(.x, "_")[[1]] %>% head(7) %>% tail(1)) %>% unlist,
         threshold_method = map(file_path, ~str_split(.x, "_")[[1]] %>% head(4) %>% tail(1)) %>% unlist) %>%
  select(-file_path) %>%
  unnest
```

Remove accidently added UDG lib control

```{r}
cumperdec <- cumperdec %>% filter(sample != "LIB025.A0301")
```


Load metadata

```{r}
metadata <- read_tsv(paste0(proj_dir, "00-documentation.backup/02-calculus_microbiome-deep_evolution-individualscontrolssources_metadata_20190523.tsv"))
```

Set ordering

```{r}
env_ordering <- c("Howler_Monkey",
                                         "Gorilla_1",
                                         "Gorilla_2",
                                         "Gorilla_3",
                                         "Chimp_1",
                                         "Chimp_2",
                                         "Chimp_3",
                                         "Chimp_4",
                                         "Neanderthal",
                                         "PreagriculturalHuman_1",
                                         "PreagriculturalHuman_2",
                                         "PreantibioticHuman_1",
                                         "PreantibioticHuman_2",
                                         "ModernDayHuman_1",
                                         "ModernDayHuman_2",
                                         "ExtractionControl",
                                         "LibraryControl",
                                         "subPlaque",
                                         "supPlaque",
                                         "urbanGut",
                                         "ruralGut",
                                         "skin",
                                         "EnvironmentalControl",
                                         "sediment")


hostgeneral_ordering <- c("Howler" , "Gorilla", "Chimp", "Neanderthal", 
                          "PreagriculturalHuman", "PreantibioticHuman", 
                          "ModernDayHuman",  "ExtractionControl", 
                          "LibraryControl", "plaque", "gut", "skin", 
                          "EnvironmentalControl", "sediment")

```


Bind together

```{r}
final_data <- left_join(cumperdec, metadata, by = c("sample" = "#SampleID")) %>%
  mutate(Env = factor(Env, levels = env_ordering),
         Host_General = factor(Host_General, levels = hostgeneral_ordering)) %>%
  select(database, threshold, threshold_method, sample, withinfluctuationvariation_pass, 
         Env, Host_General, Host_Genus, Study, Type)
  
```

# Data Summarising

```{r}
summary_env <- final_data %>%
  group_by(database, Env) %>%
  filter(Type == "Calculus") %>%
  summarise(Passed = sum(withinfluctuationvariation_pass),
            Failed = sum(!withinfluctuationvariation_pass)) %>%
  mutate(Total_Samples = Passed + Failed,
         Percent_Failed = round(Passed / Total_Samples * 100, 2))

summary_env

summary_general <- final_data %>%
  group_by(database, Host_General) %>%
  filter(Type == "Calculus") %>%
  summarise(Passed = sum(withinfluctuationvariation_pass),
            Failed = sum(!withinfluctuationvariation_pass)) %>%
  mutate(Total_Samples = Passed + Failed,
         Percent_Passed = round(Passed / Total_Samples * 100, 2)) %>%
  rename(Database = database,
         `Host Group` = Host_General, 
         `Total Samples (n)` = Total_Samples,
         `Passed (%)` = Percent_Passed)

summary_general
```

To save the three

```{r}
write_tsv(final_data, paste0("/home/fellows/Dropbox (MPI SHH)/Projects/0-Calculus_Oral_Evolution/1-Oral_Deep_Evolution/06-manuscripts/SupplementaryData/SuppDataSX-PassingCumulativeDecay/SuppDataSX-PassCumulativeDecay_PerIndividualPassFail_", format(Sys.Date(), "%Y%m%d"), ".tsv"))
#write_tsv(summary_env, paste0("/home/fellows/Dropbox (MPI SHH)/Projects/0-Calculus_Oral_Evolution/1-Oral_Deep_Evolution/06-manuscripts/SupplementaryTables/SuppTableSX-PassingCumulativeDecay/SuppDataSX-PassCumulativeDecay_EnvPassFailSummary_", format(Sys.Date(), "%Y%m%d"), ".tsv"))
write_tsv(summary_general, paste0("/home/fellows/Dropbox (MPI SHH)/Projects/0-Calculus_Oral_Evolution/1-Oral_Deep_Evolution/06-manuscripts/SupplementaryTables/SuppTableSX-PassingCumulativeDecay/SuppDataSX-PassCumulativeDecay_GeneralPassFailSummary_", format(Sys.Date(), "%Y%m%d"), ".tsv"))

```

