---
title: "Overlap of Amino acid, Carbohydrate, and Lipid protein/orthologs between SEED and KEGG"
output: html_notebook
---
# Preparation

Start by using decontam to remove the SEED categories that appear to come from contaminants, using the program decontam. 
First load the libraries you'll use
```{r}
library("data.table")
library(janitor)
library(tidyverse)
```

Set the working directory of the notebook to the directory of the notebook itself.  

```{r setup}
knitr::opts_knit$set(root.dir = normalizePath("."))
```

Load in the files that were exported from 144-imv-oral_evolution_humann2_fxn.Rmd and 148-imv-aadder_evolution_function.Rmd
```{r read_tables}
aminoacids.kegg <- fread("../04-analysis/screening/humann2/tables/aminoacids.kegg.tsv")
carbs.kegg <- fread("../04-analysis/screening/humann2/tables/carbs.kegg.tsv")
lipids.kegg <- fread("../04-analysis/screening/humann2/tables/lipids.kegg.tsv")

aminoacids.seed <- read.table("../04-analysis/screening/humann2/tables/aminoacids.seed.tsv", header = T)
carbs.seed <- read.table("../04-analysis/screening/humann2/tables/carbs.seed.tsv", header = T)
fattyacids.seed <- read.table("../04-analysis/screening/humann2/tables/fattyacids.seed.tsv", header = T)

```

Make the EC numbers a new column
```{r ec_columns_kegg}
# for amino acids
aminoacids.kegg <- aminoacids.kegg %>%
  mutate(EC_number = sapply(Enzyme_name, function(f) {
  unlist(str_split(f, "\\[EC:"))[2]
  }))
aminoacids.kegg$EC_number <- gsub("\\]","", aminoacids.kegg$EC_number)

aminoacids.kegg <- aminoacids.kegg %>%
mutate(EC_number_1 = sapply(EC_number, function(f) {
  unlist(str_split(f, " "))[1]
  })) %>%
mutate(EC_number_2 = sapply(EC_number, function(f) {
  unlist(str_split(f, " "))[2]
  })) %>%
  mutate(EC_number_3 = sapply(EC_number, function(f) {
  unlist(str_split(f, " "))[3]
  }))

# for carbohydrates
carbs.kegg <- carbs.kegg %>%
  mutate(EC_number = sapply(Enzyme_name, function(f) {
  unlist(str_split(f, "\\[EC:"))[2]
  }))
carbs.kegg$EC_number <- gsub("\\]","", carbs.kegg$EC_number)

carbs.kegg <- carbs.kegg %>%
mutate(EC_number_1 = sapply(EC_number, function(f) {
  unlist(str_split(f, " "))[1]
  })) %>%
mutate(EC_number_2 = sapply(EC_number, function(f) {
  unlist(str_split(f, " "))[2]
  })) %>%
  mutate(EC_number_3 = sapply(EC_number, function(f) {
  unlist(str_split(f, " "))[3]
  }))

# for lipids
lipids.kegg <- lipids.kegg %>%
  mutate(EC_number = sapply(Enzyme_name, function(f) {
  unlist(str_split(f, "\\[EC:"))[2]
  }))
lipids.kegg$EC_number <- gsub("\\]","", lipids.kegg$EC_number)

lipids.kegg <- lipids.kegg %>%
mutate(EC_number_1 = sapply(EC_number, function(f) {
  unlist(str_split(f, " "))[1]
  })) %>%
mutate(EC_number_2 = sapply(EC_number, function(f) {
  unlist(str_split(f, " "))[2]
  })) %>%
  mutate(EC_number_3 = sapply(EC_number, function(f) {
  unlist(str_split(f, " "))[3]
  }))

```

```{r ec_columns_seed}
aminoacids.seed <- aminoacids.seed %>%
  mutate(EC_number = sapply(protein, function(f) {
  unlist(str_split(f, "\\(EC "))[2]
  }))
aminoacids.seed$EC_number <- gsub("\\)","", aminoacids.seed$EC_number)
aminoacids.seed$EC_number <- gsub(", SufS subfamily","", aminoacids.seed$EC_number)

carbs.seed <- carbs.seed %>%
  mutate(EC_number = sapply(protein, function(f) {
  unlist(str_split(f, "\\(EC "))[2]
  }))
carbs.seed$EC_number <- gsub("\\)","", carbs.seed$EC_number)

fattyacids.seed <- fattyacids.seed %>%
  mutate(EC_number = sapply(protein, function(f) {
  unlist(str_split(f, "\\(EC "))[2]
  }))
fattyacids.seed$EC_number <- gsub("\\)","", fattyacids.seed$EC_number)

```

```{r ec_list_kegg}
aa_EC_kegg <- aminoacids.kegg %>%
  select(EC_number_1)

aa_EC_kegg <- bind_rows(aa_EC_kegg, aminoacids.kegg %>%
  select(EC_number_2) %>%
    rename(EC_number_1 = EC_number_2))

aa_EC_kegg <- bind_rows(aa_EC_kegg, aminoacids.kegg %>%
  select(EC_number_3) %>%
    rename(EC_number_1 = EC_number_3)) %>%
   drop_na() %>%
  rename(EC_number = EC_number_1) %>%
  distinct(EC_number)


carb_EC_kegg <- carbs.kegg %>%
  select(EC_number_1)

carb_EC_kegg <- bind_rows(carb_EC_kegg, carbs.kegg %>%
  select(EC_number_2) %>%
    rename(EC_number_1 = EC_number_2))

carb_EC_kegg <- bind_rows(carb_EC_kegg, carbs.kegg %>%
  select(EC_number_3) %>%
    rename(EC_number_1 = EC_number_3)) %>%
  drop_na() %>%
  rename(EC_number = EC_number_1) %>%
  distinct(EC_number)


lipids_EC_kegg <- lipids.kegg %>%
  select(EC_number_1)

lipids_EC_kegg <- bind_rows(lipids_EC_kegg, lipids.kegg %>%
  select(EC_number_2) %>%
    rename(EC_number_1 = EC_number_2))

lipids_EC_kegg <- bind_rows(lipids_EC_kegg, lipids.kegg %>%
  select(EC_number_3) %>%
    rename(EC_number_1 = EC_number_3)) %>%
  drop_na() %>%
  rename(EC_number = EC_number_1) %>%
  distinct(EC_number)

```

```{r ec_matching}
EC_aminoacids <- full_join(aminoacids.seed, aa_EC_kegg)
EC_carbohydrates <- full_join(carbs.seed, carb_EC_kegg)
EC_lipids <- full_join(fattyacids.seed, lipids_EC_kegg)

EC_aminoacids_shared <- inner_join(aminoacids.seed, aa_EC_kegg)
EC_carbohydrates_shared <- inner_join(carbs.seed, carb_EC_kegg)
EC_lipids_shared <- inner_join(fattyacids.seed, lipids_EC_kegg)

```










